### 2022-01-17 manuscript/field_data.tex
revisao geophysics inclusas

We applied the convolutional equivalent layer method to the aeromagnetic data of Carajás, 
northern Brazil.
The survey is composed of $131$ flight lines along north-south direction with line spacing of 
$\Delta y = 3,000$ m. 
Data were measured with spacing $\Delta x = 7.65$ m along lines, with an average distance 
to the ground of $900$ m. % (TEM QUE CHECAR)
The total number of observation points is $N = 6,081,345$. Figure \ref{fig:carajas_real_data_mag} 
shows the observed total-field anomaly data over the study area.

We compare the results obtained with an interpolated regular grid of $10,000 \times 131$ points, 
by using the nearest neighbor algorithm, and a decimated irregular grid, also with $10,000 \times 131$
points, totaling $N = 1,310,000$ observation points in both cases. 
Both application were made with an Intel core i7 7700HQ@2.8GHz processor in single-processing and 
single-threading modes. 
Figures \ref{fig:carajas_real_data_decimated_gridline}a and 
\ref{fig:carajas_real_data_decimated_gridline}b show, respectively, the data obtained by interpolation
and decimation. 
With $1,310,000$ observation points, it would be necessary $12.49$ Terabytes of RAM to store the full
sensitivity matrix with the classical method. 
In this case, our method uses only $59.97$ Megabytes, allowing regular desktop computers to be able 
to process this amount of data.

As the study area is very large, the main magnetic field varies with position.
For this application, we set the main field direction as that of a mid location 
(latitude $-6.5^{\circ}$ and longitude $-50.75^{\circ}$) where the declination is $-19.86^{\circ}$ and
the inclination is $-7.4391^{\circ}$. Both values were calculated using the magnetic field calculator from NOAA
at 1st January, 2014 (epoch of the survey). 
%
%\textbf{Alternative text} $\rightarrow$
%
%As the study area is very large, the main magnetic field varies with position.
%For this application, we set the main field direction as that of a mid location 
%(latitude $-6.5^{\circ}$ and longitude $-50.75^{\circ}$) where the declination is $-19.86^{\circ}$ 
%and inclination is $XXXXXX^{\circ}$ according to IGRF model at 1st January, 2014 (epoch of the survey).
%(É ESTRANHO CALCULAR A DECLINAÇÃO COM UM MODELO E A INCLINAÇÃO COM OUTRO)
%
%$\leftarrow$ \textbf{Alternative text}
%
We set the equivalent layer at $300$ meters above the ground ($600$ m below the data).
By applying our method to the interpolated regular grid 
(Figure \ref{fig:carajas_real_data_decimated_gridline}a), we obtain the  predicted data shown in 
Figure \ref{fig:carajas_gz_predito_mag_gridline}a and data residuals 
(Figure \ref{fig:carajas_gz_predito_mag_gridline}b) with mean $0.0762$ nT and the standard deviation 
$0.4886$ nT, revealing an acceptable data fitting.
Our method took $\approx 390.80$ seconds to converge at about $200$ iterations. % (TEM QUE CHECAR)
%
%POR QUE A CURVA DE CONVERGENCIA NÃO FOI MOSTRADA?
%
By applying our method to the decimated irregular grid 
(Figure \ref{fig:carajas_real_data_decimated_gridline}b), we obtain the predicted data shown in 
Figure \ref{fig:carajas_gz_predito_mag_decimated}a and data residuals 
(Figure \ref{fig:carajas_gz_predito_mag_decimated}b) with mean $0.0717$ nT and standard deviation 
$0.3144$ nT. In this case, our method took $\approx 385.56$ seconds 
to converge at about $200$ iterations (Figure \ref{fig:convergence_carajas_mag_decimated}).
%(TEM ALGO ESTRANHO AQUI. COMO QUE 2000 ITERAÇÕES FOI MAIS MAIS RÁPIDO DO QUE AS 200 ITERAÇÕES DO GRID INTERPOLADO?).
The convergence curve reveals a good convergence rate obtained with the decimated 
irregular grid. This result shows the robustness of our method in processing irregular grid.

@@

Finally, Figure \ref{fig:up5000_carajas_decimated_mag} shows the upward-continued magnetic data to a
horizontal plane located at $5, \,000$ m using the estimated equivalent layer obtained by applying our
method to the decimated irregular grid (Figure \ref{fig:carajas_real_data_decimated_gridline}b).
This process took $\approx 2.64$ seconds, showing good results without visible errors or border 
effects.

>>>>

We applied the convolutional equivalent layer method to the aeromagnetic data of Carajás, 
northern Brazil.
The survey is composed of $131$ flight lines along north-south direction with line spacing of 
$\Delta y = 3,000$ m. 
Data were measured with average spacing $\Delta x = 7.65$ m along lines,
with an average distance to the ground of $900$ m. 
The total number of observation points is $N = 6,081,345$. Figure \ref{fig:carajas_residuals_comparison}a
shows the observed total-field anomaly data over the study area.

We compare the results obtained with an interpolated regular grid of $10,000 \times 131$ points, 
by using the nearest neighbor algorithm, and a decimated irregular grid, also with $10,000 \times 131$
points. In both cases the total $N = 1,310,000$ observation points are in the original undulating surface of the flight lines. 
The decimated grid was generated by 
choosing the nearest observation points in comparison of the regular grid presented in the interpolation.
The mean and standard deviation of this irregular decimated from the regular interpolated are $6.8386$ m and $107.7343$ m in the $x$-direction and $30.8799$ m and $28.3849$ m in the $y$-direction, respectively.
Both application were made with an Intel core i7 7700HQ@2.8GHz processor in single-processing and 
single-threading modes.

As the study area is very large, the main magnetic field varies with position.
For this application, we set the main field direction as that of a mid location 
(latitude $-6.5^{\circ}$ and longitude $-50.75^{\circ}$) where the declination is $-19.86^{\circ}$ and
the inclination is $-7.4391^{\circ}$. Both values were calculated using the magnetic field calculator from NOAA
at 1st January, 2014 (epoch of the survey). We set the equivalent layer at $300$ meters above the ground ($600$ m below the data).
Figure \ref{fig:carajas_residuals_comparison}b shows the residuals obtained after using our method to fit
the interpolated data with mean $0.0762$ nT and the standard deviation  $0.4886$ nT, revealing an acceptable data fitting. Our method took $\approx 390.80$ seconds to converge at about $200$ iterations.
Figure \ref{fig:carajas_residuals_comparison}c shows the residuals obtained after using our method to fit
the decimated data with mean $0.0717$ nT and standard deviation 
$0.3144$ nT with a better fit than the produced by the interpolated data. In this case, our method took $\approx 385.56$ seconds to converge at about $200$ iterations (Figure \ref{fig:carajas_residuals_comparison}d). 
The convergence curve reveals a good convergence rate obtained with the decimated 
irregular grid. This result shows the robustness of our method in processing irregular grid. Notice that we used $200$ iterations in our method of the interpolated regular grid and the mean residual still decreasing up to $2000$. This happens because the invariance convergence criterion was met as the mean residuals are very small and each iteration decreases less than $0.00001$

With $1,310,000$ observation points, it would be necessary $12.49$ Terabytes of RAM to store the full
sensitivity matrix with the classical method. 
In this case, our method uses only $59.97$ Megabytes, allowing regular desktop computers to be able 
to process this amount of data.

%
%\textbf{Alternative text} $\rightarrow$
%
%As the study area is very large, the main magnetic field varies with position.
%For this application, we set the main field direction as that of a mid location 
%(latitude $-6.5^{\circ}$ and longitude $-50.75^{\circ}$) where the declination is $-19.86^{\circ}$ 
%and inclination is $XXXXXX^{\circ}$ according to IGRF model at 1st January, 2014 (epoch of the survey).
%(É ESTRANHO CALCULAR A DECLINAÇÃO COM UM MODELO E A INCLINAÇÃO COM OUTRO)
%
%$\leftarrow$ \textbf{Alternative text}
%

%(Figure \ref{fig:carajas_real_data_decimated_gridline}a), we obtain the  predicted data shown in 
%Figure \ref{fig:carajas_gz_predito_mag_gridline}a and data residuals 
%(Figure \ref{fig:carajas_gz_predito_mag_gridline}b) with mean $0.0762$ nT and the standard deviation 
%$0.4886$ nT, revealing an acceptable data fitting.
%Our method took $\approx 390.80$ seconds to converge at about $200$ iterations.
%
%POR QUE A CURVA DE CONVERGENCIA NÃO FOI MOSTRADA?
%

%By applying our method to the decimated irregular grid 
%(Figure \ref{fig:carajas_real_data_decimated_gridline}b), we obtain the predicted data shown in 
%Figure \ref{fig:carajas_gz_predito_mag_decimated}a and data residuals 
%(Figure \ref{fig:carajas_gz_predito_mag_decimated}b) with mean $0.0717$ nT and standard deviation 
%$0.3144$ nT. In this case, our method took $\approx 385.56$ seconds 
%to converge at about $200$ iterations (Figure \ref{fig:convergence_carajas_mag_decimated}).
%(TEM ALGO ESTRANHO AQUI. COMO QUE 2000 ITERAÇÕES FOI MAIS MAIS RÁPIDO DO QUE AS 200 ITERAÇÕES DO GRID INTERPOLADO?).

%The convergence curve reveals a good convergence rate obtained with the decimated 
%irregular grid. This result shows the robustness of our method in processing irregular grid. Notice that we used in our %method of the interpolated regular grid $200$ iterations and the mean residual still decreasing up to $2000$. This happens %because the invariance convergence criterion was met as the mean residuals are very small and each iteration decreases less %than $0.00001$

@@

Finally, Figure \ref{fig:carajas_upward_comparison}a shows the upward-continued magnetic data to a
horizontal plane located at $5, \,000$ m using the estimated equivalent layer obtained by applying our
method to the decimated irregular grid. This process took $\approx 2.64$ seconds, showing good results 
without visible errors or border effects. 
Figure \ref{fig:carajas_upward_comparison}b shows the upward-continued magnetic data to a
horizontal plane located at $5, \,000$ m using the classical Fourier filtering method to the decimated irregular grid.
This process took $\approx 0.5$ seconds. Altough, no border effect could be observed, this result showed unexpected high values instead of the expected attenuation of the anomalies.

### 2022-01-17 manuscript/introduction.tex
revisao geophysics inclusas

Large-scale data processing with tens of thousands of data, is a reality in all areas of geophysics including the geophysical potential fields. 
The potential-field data processing includes convolution integrals which can be solved either in the space or Fourier domains.
The earliest techniques of potential-field data processing  were developed in the space domain.
For example, \cite{Peters1949} accomplished, in the space domain, the second and fourth derivatives of magnetic data and the upward- and downward-continuations of magnetic data by deriving coefficients that are
used in a  graphical convolution with the magnetic data.
Howerver, the techniques for  processing potential-field data in space-domain were soon substituted by the Fourier-domain techniques. 
\cite{Dean1958} pointed out that the operations of second derivative, analytic continuation, smoothing, the removing of residuals or regionals, and others for processing  potential-field data are
similar to the electric filter circuits in Fourier domain.
\cite{Dean1958} was the first to develop the theroy of  linear filter in Fourier domain for gravity and magnetic processing and to present filters in Fourier domain \citep[][ see Table I, p 113]{Dean1958} 
for some theoretical geophysical operations (e.g., derivatives and upward and downward continuations).
\cite{GUNN1975} presented a comprehensive analysis of processing potential-field data in Fourier domain.

An approach for processing potential-field data in space domain is the equivalent-layer technique.
The equations deductions of the equivalent layer as a solution of the Laplace's equation in the region above the source was first presented by \cite{kellogg1929} and detailed explanations can also be found in \cite{blakely1996}. 
Although the equivalent-layer technique  has been known since the 1960s in geophysical literature \citep{danes1961structure,bott1967solution,dampney1969}, its use has become feasible only recently 
because the advances in computational power.
In magnetic data procesing, some authors explored this technique for calculating the first and second vertical derivatives fields \citep{emilia1973}, reduction to the pole \citep{silva1986,oliveirajr-etal2013,li2014using}, upward/downward continuations \citep{hansen-miyazaki1984,li-oldenburg2010} and total magnetic induction vector components calculation \citep{sun2019constrained}.

Together with the rise in computational processing power, some works tried new implementations to increase the efficiency of the equivalent layer. 
In \cite{leao-silva1989} the authors used a shifting window over the layer, increasing the number of linear systems to be solved, but the size of each linear system is reduced. 
Another approach for a fast equivalent layer was proposed by \cite{li-oldenburg2010}  who transformed the full sensitivity matrix into a sparse one by using  the compression of the coefficient matrix 
using fast wavelet transforms based on orthonormal, compactly supported wavelets.  
\cite{oliveirajr-etal2013} divided the equivalent layer into a grid of fixed source windows.
Instead of directly calculating the the physical-property distribution of a finite set of equivalent
sources (e.g., dipoles, point of masses) arranged in the entire equivalent layer,
\cite{oliveirajr-etal2013} estimated the coefficients of a bivariate polynomial function describing 
the physical-property distribution within each equivalent-source window.
The estimated polynomial-coefficients are transformed into the physical-property distribution
and thus any standard linear transformation of the data can be performed.
Grounded on excess mass constraint, \cite{siqueira-etal2017} proposed an iterative method 
for processing large gravimetric data using the equivalent layer without requiring the solution 
of a linear system. 
In \cite{siqueira-etal2017}, the initial mass distribution over the equivalent layer is
proportional to observed gravity data and it is updated at each iteration by adding mass corrections that are proportional to the residuals of observed and estimated data.

One of the greatest obstacles to the use of the equivalent-layer technique for processing potential-field data is the solution of the associated linear system.
A wide variety of applications in mathematics and engineering that fall into Toeplitz systems propelled the development of a large variety of  methods for solving them. Direct methods were conceived by \cite{levinson1946} and by \cite{trench1964}. Currently, the iterative method of conjugate gradient is used in most cases, in \cite{chan-jin2007} the authors presented an introduction on the topic for 1D data structures of Toeplitz matrices and also for 2D data structures, which they called block-Toeplitz Toeplitz-block matrices. In both cases, the solving strategy is to embbed the Toeplitz/BTTB matrix into a Circulant/Block-Circulant Circulant-Block matrix, calculate its eigenvalues by a 1D or 2D fast Fourier transform of its first column, respectively and carry the matrix-vector product between kernel and parameters at each iteration of the conjugate gradient method in a very fast manner.

@@

In this work, the convolutional equivalent layer using the block-Toeplitz Toeplitz-block idea, presented in \cite{takahashi2020convolutional}, will be used to solve the linear system required to estimate the physical property that produces a magnetic field on regular grids. 
Here, we achive very fast solutions using a conjugate gradient algorithm combined with the fast Fourier transform. We present a novel method of exploring the symmetric structures of the second derivatives of the inverse of the distance contained in the magnetic kernel, to keep the memory RAM usage to the minimal by using only one equivalent source to carry the calculations of the forward problem. We also show tests of the magnetic convolutional equivalent layer when irregular grids are used. The convergence of the conjugate gradient maintains in an acceptable level even using irregular grids. 
Our results show the good performance of our method in producing fast and robust solutions for processing large amounts of magnetic data using the equivalent layer technique.

>>>>

Large-scale data processing with tens of thousands of data is a reality in all areas of geophysics including the geophysical potential fields. 
The potential-field data processing includes convolution integrals which can be solved either in the space or Fourier domains.
The earliest techniques of potential-field data processing  were developed in the space domain.
For example, \cite{Peters1949} accomplished, in the space domain, the second and fourth derivatives of magnetic data and the upward- and downward-continuations of magnetic data by deriving coefficients that are
used in a  graphical convolution with the magnetic data.
However, the techniques for  processing potential-field data in space-domain were soon substituted by the Fourier-domain techniques. 
\cite{Dean1958} pointed out that the operations of second order derivatives, analytic continuation, smoothing, the removing of residuals or regionals, and others for processing  potential-field data are
similar to the electric filter circuits in Fourier domain.
\cite{Dean1958} was the first to develop the theory of linear filters in Fourier domain for gravity and magnetic processing and to present filters in Fourier domain \citep[][ see Table I, p 113]{Dean1958} 
for some theoretical geophysical operations (e.g., derivatives and upward and downward continuations).
\cite{GUNN1975} presented a comprehensive analysis of processing potential-field data in Fourier domain.

An approach for processing potential-field data in space domain is the equivalent-layer technique.
The equations deductions of the equivalent layer as a solution of the Laplace's equation in the region above the source was first presented by \cite{kellogg1929} and detailed explanations can also be found in \cite{blakely1996}. 
Although the equivalent-layer technique  has been known since the 1960s in geophysical literature \citep{danes1961structure,bott1967solution,dampney1969}, its use has become feasible only recently 
because of the advances in computational power.
In magnetic data procesing, some authors explored this technique for calculating the first and second vertical derivatives of the fields \citep{emilia1973}, reduction to the pole \citep{silva1986,oliveirajr-etal2013,li2014using}, upward/downward continuations \citep{hansen-miyazaki1984,li-oldenburg2010} and total magnetic induction vector components calculation \citep{sun2019constrained}.

Together with the rise in computational processing power, some works tried new implementations to increase the efficiency of the equivalent layer. 
In \cite{leao-silva1989} the authors used a shifting window over the layer to increase the number of linear systems to be solved, but to reduced the size of each linear system at the same time. 
Another approach for a fast equivalent layer was proposed by \cite{li-oldenburg2010}  who transformed the full sensitivity matrix into a sparse one by using  the compression of the coefficient matrix 
using fast wavelet transforms based on orthonormal, compactly supported wavelets.  
\cite{oliveirajr-etal2013} divided the equivalent layer into a grid of fixed source windows.
Instead of directly calculating the the physical-property distribution of a finite set of equivalent
sources (e.g., dipoles, point of masses) arranged in the entire equivalent layer,
\cite{oliveirajr-etal2013} estimated the coefficients of a bivariate polynomial function describing 
the physical-property distribution within each equivalent-source window.
The estimated polynomial-coefficients are transformed into the physical-property distribution
and thus any standard linear transformation of the data can be performed.
Grounded on excess mass constraint, \cite{siqueira-etal2017} proposed an iterative method 
for processing large gravimetric data using the equivalent layer without requiring the solution 
of a linear system. 
In \cite{siqueira-etal2017}, the initial mass distribution over the equivalent layer is
proportional to observed gravity data and it is updated at each iteration by adding mass corrections that are proportional to the residuals of observed and estimated data.

One of the greatest obstacles to the use of the equivalent-layer technique for processing potential-field data is the solution of the associated linear system.
A wide variety of applications in mathematics and engineering that fall into Toeplitz systems propelled the development of a large variety of  methods for solving them. Direct methods were conceived by \cite{levinson1946} and by \cite{trench1964}. Currently, the iterative method of conjugate gradient is used in most cases, in \cite{chan-jin2007} the authors presented an introduction on the topic for 1D data structures of Toeplitz matrices and also for 2D data structures, which they called block-Toeplitz Toeplitz-block matrices. In both cases, the solving strategy is to embed the Toeplitz/BTTB matrix into a Circulant/Block-Circulant Circulant-Block matrix, calculate its eigenvalues by a 1D or 2D fast Fourier transform of its first column, respectively and carry the matrix-vector product between kernel and parameters at each iteration of the conjugate gradient method in a very fast manner.

@@

In this work, the convolutional equivalent layer using the block-Toeplitz Toeplitz-block idea, presented in \cite{takahashi2020convolutional}, will be used to solve the linear system required to estimate the physical property that produces a magnetic field on regular grids. 
Here, we achieve very fast solutions using a conjugate gradient algorithm combined with the fast Fourier transform. We present a novel method of exploring the symmetric structures of the second derivatives of the inverse of the distance contained in the magnetic kernel, to keep the memory RAM usage to the minimal by using only one equivalent source to carry the calculations of the forward problem. We also show tests of the magnetic convolutional equivalent layer when irregular grids are used. The convergence of the conjugate gradient maintains in an acceptable level even using irregular grids. 
Our results show the good performance of our method in producing fast and robust solutions for processing large amounts of magnetic data using the equivalent layer technique.
